\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{url}
\usepackage{xspace}
\usepackage{amsmath}

\newcommand{\method}{SchemaShield-Lite\xspace}

\begin{document}

\title{Inference-Time Repair for Robust Tool Calling Under Interface Drift}
\titlerunning{Inference-Time Repair for Robust Tool Calling}

\author{Anonymous Authors}
\authorrunning{Anonymous Authors}
\institute{Draft prepared for NGEN-AI 2026}

\maketitle

\begin{abstract}
Tool-calling language models rely heavily on surface-form tool interfaces, including
descriptions, parameter names, and candidate tool sets. In realistic deployments,
these interfaces drift after release even when the underlying tool semantics remain
unchanged. We study this problem as \emph{interface drift} and evaluate a lightweight
training-free mitigation, \method, built from canonical tool cards, validation, and
one-shot repair. We evaluate four open models across two benchmarks (DICE-BENCH at
$n{=}300$ and BFCL at $n{=}200$) under three drift perturbations applied individually
and in combination. On DICE-BENCH with Qwen3.5-9B, accuracy drops from $0.750$ to
$0.713$ under drift, while \method recovers performance to $0.827$ and fixes
$17/18$ drift-induced failures with zero harms. Across all four models, repair
recovers drift-induced failures with zero harms. A naive-retry baseline scores
$0.700$ on DICE (worse than drifted) and $0.650$ on BFCL, demonstrating that the
gains come from structured validation feedback rather than from a second attempt.
Drift-type ablation shows that candidate-set drift dominates: description and
schema drift alone cause $< 1.5$ point degradations on modern models, while
distractor tools cause most of the observed harm. A component ablation reveals
that the canonical tool card is not the load-bearing component: validation feedback
combined with forced-tool repair accounts for essentially all of the recovery. On
BFCL with Llama-4-Scout, \method improves accuracy from $0.685$ to $0.855$, a
$17.0$ point lift over drifted performance.
\keywords{tool calling \and language models \and robustness \and interface drift \and repair}
\end{abstract}

\section{Introduction}
Modern language-model agents increasingly depend on external tools for retrieval,
execution, planning, and actuation. In practice, however, tool use is mediated
entirely through text: tool descriptions, parameter schemas, enum values, and
candidate tool lists. Recent work shows that these interface details strongly affect
tool selection and argument generation
even when the tool semantics are unchanged \cite{guo2026rewrite,faghih2025gaming}.
This creates a deployment problem: tool interfaces drift over time as developers
rename parameters, revise descriptions, add examples, or expand candidate tool sets.

We study this problem as \emph{interface drift}: benign, semantics-preserving
changes to tool interfaces that occur after deployment. Our goal is not to train a
new tool-calling model, but to quantify drift sensitivity and test whether a compact
inference-time repair layer can recover reliability without retraining.

We instantiate drift with three realistic perturbations: (i) stale description
examples that mention legacy field names, (ii) schema renaming, and (iii) distractor
tools. We evaluate four open models (Qwen3.5-9B, Qwen3.5-35B-A3B, Llama-4-Scout, and
Mistral-Small-24B) on a 300-example DICE-BENCH slice \cite{jang2025dicebench} and a
200-example BFCL slice \cite{bfcl2024}. We apply \method, a deterministic
inference-time stack combining canonical tool rendering, validation, and one-shot
forced-tool repair.

This paper makes four empirical contributions that substantially revise the
narrative of prior interface-drift work:

\begin{enumerate}
\item \textbf{Inference-time repair recovers drift-induced failures across models
and benchmarks with zero harms.} On DICE-BENCH, \method recovers $17/18$
drift-induced failures on Qwen3.5-9B and achieves zero repair harms across all four
models. On BFCL, repair improves Llama-4-Scout by $17$ points and Qwen3.5-9B by
$15.5$ points over drifted performance.

\item \textbf{A naive-retry baseline underperforms even the drifted model.} On
DICE-300, naive retry scores $0.700$ versus drifted $0.713$; on BFCL, $0.650$ vs
$0.640$. This rules out the trivial explanation that repair works simply by giving
the model a second attempt.

\item \textbf{Candidate-set drift is the dominant failure mode on modern models.}
Under description-only drift or schema-only drift, Qwen3.5-9B accuracy drops by just
$0.013$ and $0.007$ points respectively. Adding distractor tools causes a $0.047$
drop, matching the full combined drift.

\item \textbf{Validation feedback, not the canonical tool card, is the load-bearing
component of \method.} A component ablation shows that \emph{validation-plus-retry
without the canonical card} recovers $16$ of $17$ drift-induced failures, nearly
matching the full pipeline. The canonical card alone never triggers repair and
recovers nothing.
\end{enumerate}

\section{Related Work}
Recent work has established that tool interfaces are not neutral wrappers around
tool use; they are part of the learning and inference problem itself.
Guo et al.~\cite{guo2026rewrite} show that rewriting tool descriptions can
substantially improve agent reliability, especially when candidate tool sets are
large. Faghih et al.~\cite{faghih2025gaming} show the opposite direction:
small edits to tool descriptions can dramatically shift tool preference, exposing a
fragility in description-mediated tool use.

Several recent papers study robustness and recovery in tool-use systems.
Zhang et al.~\cite{zhang2026fission} show that execution-error recovery remains a
major weakness, particularly for smaller models, and propose RL-based training to
improve it. Our setting is narrower and more deployment-oriented: we target
inference-time robustness under interface drift rather than post-training
optimization.

Benchmark work has also highlighted that realistic function calling is still far from
solved. ComplexFuncBench introduces multi-step and constrained long-context function
calling \cite{zhong2025complexfuncbench}, while DICE-BENCH focuses on multi-round,
multi-party dialogues with tool-related information dispersed across conversation
turns \cite{jang2025dicebench}. We evaluate on both DICE-BENCH and the Berkeley
Function Calling Leaderboard (BFCL) \cite{bfcl2024} because they capture
complementary settings: DICE-BENCH exercises conversational multi-turn dialogues,
while BFCL tests single-turn tool calls across diverse APIs.

Compared with dataset-generation work such as APIGen \cite{liu2024apigen}, our
objective is different. We do not create new large-scale tool-calling training data.
Instead, we evaluate whether a lightweight repair layer can stabilize existing models
against a specific class of post-deployment failures.

\section{Problem Formulation}
Let a tool be represented by an interface
$I = (d, s, C)$, where $d$ is the natural-language description, $s$ is the argument
schema, and $C$ is the candidate tool set shown to the model. The underlying tool
semantics are assumed fixed. An \emph{interface drift} operator $\Delta$ maps $I$ to
a new interface $\Delta(I)$ that preserves tool semantics but perturbs surface form.

We study three drift types:
\begin{itemize}
\item \textbf{Description drift}: add stale or legacy examples to the description that
refer to older field names.
\item \textbf{Schema drift}: rename parameters while preserving semantics.
\item \textbf{Candidate-set drift}: add distractor tools with overlapping descriptions.
\end{itemize}

Given user context $x$, tool interface $I$, and model $M$, the base system outputs a
tool call $y = M(x, I)$. We evaluate three stages:
\begin{enumerate}
\item \textbf{Original}: inference on the original tool interface.
\item \textbf{Drifted}: inference on the drifted interface $\Delta(I)$.
\item \textbf{Repaired}: inference-time repair of an invalid or incomplete drifted call.
\end{enumerate}

Our central question is whether an inference-time repair layer can recover the
performance lost from Original to Drifted without harming examples that were already
correct.

\section{Method}
\subsection{\method}
\method is a lightweight inference-time robustness layer with three parts:

\paragraph{Canonical tool card.}
Each tool is rendered into a compact canonical form containing the tool name,
description, required fields, and field-level type information. This normalizes
interface presentation across tools and emphasizes current field names.

\paragraph{Validation.}
After the base model emits a tool call, a validator checks tool name, required
fields, and argument structure. Validation produces a structured error report rather
than a binary failure signal.

\paragraph{One-shot repair.}
If the drifted call is invalid, the model receives a repair prompt containing the
task context, canonical tool card, invalid call, and validation errors. Repair uses a
single constrained call. In the strongest configuration we use
\emph{forced-tool repair}: the repair step is constrained to the intended tool when
provider support allows it, with JSON fallback if the provider does not support
named tool choice.

As our component ablation in Section~\ref{sec:component-ablation} shows, it is the
combination of structured validation errors and forced-tool repair that does the
heavy lifting. The canonical card is useful at the initial prediction step as
drift-prevention prompting, but contributes negligibly to repair-stage recovery.

\subsection{Repair Prompting Details}
Two implementation details turned out to matter materially:
\begin{itemize}
\item The repair prompt explicitly instructs the model to infer all required fields
from the dialogue even when the invalid call is empty.
\item The repair stage uses a larger token budget than the base prediction stage.
Raising repair capacity from the earlier default to $512$ tokens fixed truncated
repair outputs on longer DICE dialogues.
\end{itemize}

\section{Experimental Setup}
\subsection{Benchmarks and Data Slices}
We evaluate on two benchmarks. For DICE-BENCH \cite{jang2025dicebench}, we use a
300-example subset drawn from single-function round-1 dialogues (allowing duplicate
functions to reach sufficient scale). For BFCL \cite{bfcl2024}, we use a 200-example
subset drawn in equal parts from four categories: simple Python, multiple-function,
live-simple, and live-multiple.

\subsection{Models}
We evaluate four open instruction-tuned models accessed via OpenRouter:

\begin{itemize}
\item \textbf{Qwen3.5-9B}: our primary model.
\item \textbf{Qwen3.5-35B-A3B}: larger MoE variant.
\item \textbf{Llama-4-Scout}: Meta's 2025 open model.
\item \textbf{Mistral-Small-3.2-24B}: recent Mistral dense model.
\end{itemize}

All experiments use temperature $0.0$ for deterministic-style evaluation.

\subsection{Drift Configurations}
For DICE main runs, we use \emph{mild} severity: two description modes
(\texttt{legacy\_example}, \texttt{marketing}), two schema modes
(\texttt{rename\_parameters}, \texttt{reorder\_parameters}), and one distractor
tool. For BFCL we use \emph{medium} severity: three description modes, three schema
modes, and three distractors, reflecting a more aggressive drift scenario.

\subsection{Metrics}
We report:
\begin{itemize}
\item \textbf{Original score}: accuracy on the unmodified interface.
\item \textbf{Drifted score}: accuracy under interface drift.
\item \textbf{Repaired score}: accuracy after one-shot repair.
\item \textbf{Naive retry score}: accuracy under a second identical drifted call
(no canonical card, no validation feedback). This baseline isolates the value of
simply retrying.
\item \textbf{Originally-correct slice}: the subset of examples that the model solves
correctly before drift. This slice isolates \emph{true drift harm} from baseline
errors.
\item \textbf{Repair recoveries}: drift-induced failures recovered by repair.
\item \textbf{Repair harms}: drifted-correct examples broken by repair.
\item \textbf{Repair trigger rate}: fraction of examples for which the validator
invokes the repair step.
\end{itemize}

\section{Results}
\subsection{Main Results Across Models and Benchmarks}
Table~\ref{tab:main-results} summarizes results across all four models on DICE-300
and both models evaluated on BFCL-200. Across every model and both benchmarks,
\method improves over the drifted baseline with \emph{zero repair harms}.

\begin{table}[t]
\caption{Main results on DICE-300 and BFCL-200. OC = originally-correct count;
Rec = recoveries; Harms = repair harms on originally-correct slice.}
\label{tab:main-results}
\centering
\small
\begin{tabular}{llcccccccc}
\toprule
Benchmark & Model & $n$ & Orig. & Drift. & Repair & OC & Rec/Miss & Harms & Trig\% \\
\midrule
\multirow{4}{*}{DICE-300}
& Qwen3.5-9B      & 300 & 0.750 & 0.713 & \textbf{0.827} & 225 & 17/18 & 0 & 21.0 \\
& Qwen3.5-35B-A3B & 300 & 0.773 & 0.713 & \textbf{0.820} & 232 & 20/25 & 0 & 21.3 \\
& Llama-4-Scout   & 300 & 0.707 & 0.717 & \textbf{0.807} & 212 & 12/16 & 0 & 13.3 \\
& Mistral-24B     & 300 & 0.740 & 0.777 & \textbf{0.827} & 222 & 8/12  & 0 & 9.0  \\
\midrule
\multirow{2}{*}{BFCL-200}
& Qwen3.5-9B      & 200 & 0.790 & 0.640 & \textbf{0.795} & 158 & 23/35 & 0 & 31.0 \\
& Llama-4-Scout   & 200 & 0.535 & 0.685 & \textbf{0.855} & 107 & 9/15  & 0 & 21.0 \\
\bottomrule
\end{tabular}
\end{table}

The BFCL-200 configuration uses medium-severity drift and shows substantially larger
effects than DICE mild drift. On Qwen3.5-9B, drift drops accuracy by $15.0$ points
($0.790 \to 0.640$) and repair fully recovers ($0.795$). On Llama-4-Scout, repair
improves accuracy by $17.0$ points over drifted and $32.0$ points over original.

\subsection{Naive Retry Is Not a Viable Baseline}
\label{sec:naive-retry}
A first-order alternative explanation is that repair works simply because the model
gets a second attempt. We test this by running a \emph{naive retry} call that uses
the same drifted tools but without the canonical card or the validation error
report. Table~\ref{tab:naive-retry} shows that naive retry underperforms even the
drifted baseline: $0.700$ vs $0.713$ on DICE and $0.650$ vs $0.640$ on BFCL. In
contrast, \method achieves $0.827$ on DICE and $0.795$ on BFCL. The gap rules out
retry-alone as the mechanism and localizes the improvement to structured validation
feedback combined with forced-tool repair.

\begin{table}[t]
\caption{Naive retry vs \method on Qwen3.5-9B.}
\label{tab:naive-retry}
\centering
\small
\begin{tabular}{lcccc}
\toprule
Benchmark & Drifted & Naive Retry & \method & $\Delta$ over Drifted \\
\midrule
DICE-300   & 0.713 & 0.700 & \textbf{0.827} & $+0.114$ \\
BFCL-200   & 0.640 & 0.650 & \textbf{0.795} & $+0.155$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Drift-Type Ablation: Candidate-Set Drift Dominates}
\label{sec:drift-ablation}
Table~\ref{tab:drift-ablation} decomposes the drift pipeline into its three
components and evaluates each in isolation at DICE-300 on Qwen3.5-9B. Description
drift alone causes only a $0.013$ point accuracy drop. Schema drift alone causes
only $0.007$. By contrast, candidate-set drift (adding a distractor tool) causes a
$0.047$ drop---matching the full combined drift scenario ($0.050$).

\begin{table}[t]
\caption{Drift-type ablation on DICE-300 with Qwen3.5-9B.}
\label{tab:drift-ablation}
\centering
\small
\begin{tabular}{lcccr}
\toprule
Drift type & Orig. & Drift. & Repair & Drift Drop \\
\midrule
Description only     & 0.760 & 0.747 & 0.813 & $-0.013$ \\
Schema only          & 0.767 & 0.760 & 0.830 & $-0.007$ \\
Candidate-set only   & 0.757 & 0.710 & 0.830 & $-0.047$ \\
\textbf{All combined}& 0.763 & 0.713 & 0.823 & $-0.050$ \\
\bottomrule
\end{tabular}
\end{table}

This result qualifies a common framing in the interface-drift literature. For modern
instruction-tuned tool-calling models, minor description rewording and
semantics-preserving parameter renames do not meaningfully degrade accuracy---the
models are evidently robust to this class of surface perturbation. The realistic
failure mode is \emph{candidate-set confusion}: when a distractor tool with an
overlapping description is added to the candidate set, tool selection degrades.
This has implications for deployment-time monitoring: teams should prioritize
tracking tool-catalog growth over monitoring description or schema revisions.

\subsection{Component Ablation: Validation, Not Canonical Card, Drives Recovery}
\label{sec:component-ablation}
Table~\ref{tab:component-ablation} evaluates three variants of \method on DICE-300
with Qwen3.5-9B: (i) canonical card only, where the card is prepended to the initial
prompt but no validation/repair is applied; (ii) validation-plus-retry, which uses
validation errors in a repair prompt but omits the canonical tool card from the
repair context; and (iii) full \method (card + validation + repair).

\begin{table}[t]
\caption{Component ablation on DICE-300 with Qwen3.5-9B.}
\label{tab:component-ablation}
\centering
\small
\begin{tabular}{lcccccc}
\toprule
Variant & Drift. & Repair & Rec/Miss & Harms & Trig\% \\
\midrule
Canonical card only            & 0.770 & 0.770 & 0/9   & 0 & 0.0 \\
Validation + retry (no card)   & 0.730 & 0.830 & 16/17 & 0 & 20.0 \\
\textbf{Full \method}          & 0.713 & \textbf{0.827} & 17/18 & 0 & 21.0 \\
\bottomrule
\end{tabular}
\end{table}

Three observations. First, prepending the canonical card at the initial prompt
\emph{reduces} drift damage ($0.770$ drifted vs $0.713$ for the base drifted run),
indicating that the card is useful as drift-prevention prompting---it helps the
model resist drift at generation time. Second, however, canonical card alone never
triggers the validator and therefore recovers zero drift-induced failures. Third,
validation-plus-retry without the canonical card achieves $0.830$ repaired accuracy
and $16/17$ recoveries, essentially matching the full pipeline ($0.827$, $17/18$).

The practical implication is that the load-bearing component of \method is the
validation error feedback loop combined with forced-tool repair, not the canonical
tool card. Teams implementing an inference-time robustness layer can drop the card
rendering without meaningfully hurting recovery, or keep the card only at the
initial prediction step where it provides drift-prevention benefits.

\subsection{Error Analysis}
Across the four main DICE runs, all repair harms are zero. However, not every drift
miss is recovered. On Qwen3.5-35B, repair recovers $20$ of $25$ drift misses; on
Llama-4-Scout, $12$ of $16$; on Mistral, $8$ of $12$. Inspection of the unrecovered
cases shows two recurring patterns: (i) the invalid drifted call has an empty
arguments object and the model still cannot infer all required fields from the
dialogue even with validation feedback, and (ii) for a small number of cases the
repair call selects a valid tool but fills in semantically-correct-but-not-identical
string values (for example, a paraphrased restaurant name). The former are
genuine inference failures; the latter are strict-match artifacts rather than
true drift failures.

On the main Qwen3.5-9B run, \method fixes $24$ examples that the base model got
wrong \emph{even before drift}. These are cases where the original call was
incomplete or used the wrong tool, and validation feedback plus forced-tool repair
surfaces the required fields. This is a welcome side effect: the same mechanism that
recovers drift failures also helps patch some base-model errors.

Mistral-24B exhibits an unusual pattern: its drifted score ($0.777$) is slightly
higher than its original score ($0.740$). The rewording produced by description-mode
drift may happen to resemble Mistral's instruction-tuning distribution more closely
than the unmodified DICE prompts, which is consistent with the finding that modern
models are largely robust to benign description changes.

\section{Discussion}
Our results support two narrower conclusions than the original \method framing
suggested. First, inference-time repair is a viable and deployable robustness
layer for tool-calling systems under interface drift. It works across model
families, across two benchmarks, and adds zero harms at any point. Second, the
\emph{mechanism} is validation feedback rather than canonical tool rendering,
and the \emph{target} is candidate-set drift rather than description or schema
drift.

Together these findings simplify the deployment picture. A practitioner does not
need a heavy per-tool canonicalization pipeline; an API-schema validator with a
forced-tool repair call is sufficient. And the monitoring story is clearer: the
first-order cause of post-deployment tool-calling regressions is likely tool-catalog
expansion, not description or parameter rename churn.

\section{Limitations}
Several limitations remain. First, while we evaluate on two benchmarks and four
model families, the BFCL evaluation uses only two models due to API-cost
constraints, and BFCL repair does not track token usage for the repair call itself.
Second, our drift operators focus on stale descriptions, renamed fields, and
distractors; other forms of interface drift (for example, type narrowing, enum
replacement, or multi-tool refactors) remain untested. Third, evaluation uses hosted
inference and so may incorporate provider-specific tool-calling behavior. Fourth,
the ``repair $>$ original'' effect on Qwen3.5-9B indicates that \method is also
patching baseline failures---this is useful in deployment but complicates clean
attribution of the improvement to drift recovery specifically. Finally, the
component ablation finding that the canonical card is non-essential to repair is
established on a single benchmark and model; it should be validated on BFCL and
additional models before being treated as a general rule.

\section{Conclusion}
We presented \method, a lightweight inference-time approach for robust tool calling
under interface drift. Across four open models and two benchmarks, \method recovers
drift-induced failures with zero harms: $17/18$ on Qwen3.5-9B and $20/25$ on
Qwen3.5-35B, with BFCL gains of up to $17$ points over drifted. Three findings
qualify the picture. A naive-retry baseline underperforms even the drifted model,
so repair gains cannot be explained as a second attempt. Description and schema
drift cause negligible degradation on modern models---candidate-set drift is the
dominant failure mode. And validation-plus-retry without the canonical card
nearly matches the full pipeline, localizing the load-bearing component to
structured validation feedback. Inference-time repair is therefore a practical and
deployable complement to model training for tool-mediated language systems, but its
mechanism and target are narrower and more specific than prior framing suggested.

\bibliographystyle{splncs04}
\bibliography{references}

\end{document}
